Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}

pf <- read.delim('pseudo_facebook.tsv')

qplot(x = age, y = friend_count, data = pf)

```

***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:
http://docs.ggplot2.org/current/
http://bbs.ceb-institute.org/wp-content/uploads/2011/09/handout_ggplot2.pdf
```{r ggplot Syntax}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20) +
  xlim(13,90) +
  coord_trans(y = 'sqrt')


summary(pf$age)
```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pf) + 
+   geom_point(alpha = 1/20) +
+   xlim(13,90)

ggplot(aes(x = age, y = friend_count), data = pf) + 
+   geom_jitter(alpha = 1/20) +
+   xlim(13,90)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20) +
  xlim(13,90) +
  coord_trans(y = 'sqrt')

ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) +
  xlim(13,90) +
  coord_trans(y = 'sqrt')
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
# This programming assignment
# will not be graded, but when you
# submit your code, the assignment
# will be marked as correct. By submitting
# your code, we can add to the feedback
# messages and address common mistakes
# in the Instructor Notes.

# You can assess your work by watching
# the solution video.


# Examine the relationship between
# friendships_initiated (y) and age (x)
# using the ggplot syntax.

# We recommend creating a basic scatter
# plot first to see what the distribution looks like.
# and then adjusting it by adding one layer at a time.

# What are your observations about your final plot?

# Remember to make adjustments to the breaks
# of the x-axis and to use apply alpha and jitter.

# ENTER ALL OF YOUR CODE FOR YOUR PLOT BELOW THIS LINE.
# =======================================================

ggplot(aes(x = age, y = friendships_initiated), data = pf)

ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_point(alpha = 1/20)

ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_point(alpha = 1/20) +
  xlim(13,90)

ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_point(alpha = 1/20) +
  xlim(13,90) +
  coord_trans(y = 'sqrt')

ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) +
  xlim(13,90) +
  coord_trans(y = 'sqrt')
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
library(dplyr)

age_groups <- group_by(pf, age)
pf.fc_by_age <- summarise(age_groups,
                          friend_count_mean = mean(friend_count),
                          friend_count_median = median(friend_count),
                          n = n())
head(pf.fc_by_age)

pf.fc_by_age <- arrange(pf.fc_by_age, age)

# pipeline/chain method
pf.fc_by_age2 <- pf%>%
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age)

head(pf.fc_by_age2)


```

Create your plot!

```{r Conditional Means Plot}

```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0), color = 'orange') + 
  xlim(13,90) + 
  coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .1),
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .9),
            linetype = 2, color = 'blue') +
  coord_cartesian(xlim = c(13, 90))

ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0), color = 'orange') + 
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .1),
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .9),
            linetype = 2, color = 'blue') +
  coord_cartesian(xlim = c(13, 90))


```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor.test(pf$age, pf$friend_count, method = 'pearson', alternative = 'greater')
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:
-0.027
***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(subset(pf, age <= 70), cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  geom_point()
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95))

ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_smooth(method = 'lm', color = 'red')

cor.test(pf$www_likes_received, pf$likes_received)
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}

```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
install.packages('alr3')
library(alr3)
```

Create your plot!

```{r Temp vs Month}
ggplot(aes(x = Month, y = Temp), data = Mitchell) +
+   geom_point()

ggplot(aes(x = Month, y = Temp), data = Mitchell) +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 203, 12))
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}

```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}
ggplot(aes(x=(Month%%12),y=Temp), data=Mitchell)+
  geom_point()
```

***

### A New Perspective

What do you notice?
Response: 

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age2) +
  geom_line()

# Create a new variable, 'age_with_months', in the 'pf' data frame.
# Be sure to save the variable in the data frame rather than creating
# a separate, stand-alone variable. You will need to use the variables
# 'age' and 'dob_month' to create the variable 'age_with_months'.

# Assume the reference date for calculating age is December 31, 2013.

# This programming assignment WILL BE automatically graded. For
# this exercise, you need only create the 'age_with_months' variable;
# no further processing of the data frame is necessary.

# DO NOT DELETE THIS NEXT LINE OF CODE
# ========================================================================
pf <- read.delim('/datasets/ud651/pseudo_facebook.tsv')


# ENTER YOUR CODE BELOW THIS LINE
# ========================================================================
pf$age_with_months <- pf$age + (1 - pf$dob_month/12)
```

***

### Age with Months Means

```{r Age with Months Means}
age_groups_months <- group_by(pf, age_with_months)
pf.fc_by_age_months <- summarise(age_groups_months,
                          friend_count_mean = mean(friend_count),
                          friend_count_median = median(friend_count),
                          n = n())
head(pf.fc_by_age_months)

pf.fc_by_age <- arrange(pf.fc_by_age_months, age_with_months)

# chain method
pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(pf.fc_by_age_months)

head(pf.fc_by_age_months)
```

Programming Assignment
```{r Programming Assignment}

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
ggplot(aes(x = age_with_months, y = friend_count_mean), 
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line()
  
```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}
p1 <- ggplot(aes(x = age_with_months, y = friend_count_mean), 
       data = pf.fc_by_age_months) +
  geom_line() +
  geom_smooth()

p2 <-ggplot(aes(x = age_with_months, y = friend_count_mean), 
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line() +
  geom_smooth()

grid.arrange(p2, p1, ncol = 1)
```

***

### Which Plot to Choose?
Notes:
https://simplystatistics.org/2014/02/13/loess-explained-in-a-gif/
http://en.wikipedia.org/wiki/Local_regression
***

### Analyzing Two Variables
Reflection:
http://dept.stat.lsa.umich.edu/~kshedden/Courses/Stat401/Notes/401-bivariate-slides.pdf
***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

