Lesson 3
========================================================

***

### What to Do First?
Notes:

***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
names(pf)
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
qplot(data = pf, x = dob_day, binwidth = 1) +
  scale_x_continuous(breaks=1:31)
```

***

#### What are some things that you notice about this histogram?
Response:
Higher than average numbers on 1 and 15
***

### Moira's Investigation
Notes:

***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:

#### How many of your friends do you think saw that post?
Response:

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:

```{r}
qplot(data = pf, x = dob_day) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month, ncol = 3)
```

#### Let’s take another look at our plot. What stands out to you here?
Response:
Jan 1 has a much larger number of responses than any other date
***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(data = pf, x = friend_count)
```

#### How is this plot similar to Moira's first plot?
Response:
Very heavily skewed right due to outliers between mid-300 and 5000
***

### Limiting the Axes
Notes:

```{r Limiting the Axes}
qplot(data = pf, x = friend_count, xlim = c(0, 1000))
```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(data = pf, x = friend_count, binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 25)) +
  facet_grid(gender ~ .)
```

***

### Omitting NA Values
Notes:

```{r Omitting NA Values}
qplot(data = subset(pf, !is.na(gender)), x = friend_count, binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 25)) +
  facet_grid(gender ~ .)

```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Response:

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:

***

### Tenure
Notes:

```{r Tenure}
qplot(data = pf, x = tenure/365, binwidth = .25, 
      color = I('black'), fill = I('#F79420')) + 
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7))

```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(data = pf, x = tenure/365, binwidth = .25, 
      color = I('black'), fill = I('#F79420')) + 
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7))
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(data = pf, x = tenure/365, binwidth = .25,
      xlab = 'No. of years using Facebook',
      ylab = 'No. of users in sample',
      color = I('black'), fill = I('#F79420')) + 
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7))
```

***

### User Ages
Notes:

```{r User Ages}
qplot(data = pf, x = age, binwidth = 1,
      xlab = 'Age of users',
      ylab = 'No. of users in sample',
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(limits = c(0, 115), breaks = seq(10, 113, 5))
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:
# define individual plots
p1 <- qplot(data = pf, x = friend_count, binwidth = 1)
p2 <- qplot(data = pf, x = log10(friend_count + 1), binwidth = 0.25)
p3 <- qplot(data = pf, x = sqrt(friend_count), binwidth = 1)

# define individual plots using ggplot

p1 <- ggplot(aes(x = friend_count), data = pf) + geom_histogram()
p2 <- p1 + scale_x_log10()
p3 <- p1 + scale_x_sqrt()
# arrange grid
grid.arrange(p1, p2, p3, ncol = 1)
***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
ggplot(aes(x = friend_count), data = pf) +
  geom_histogram() +
  scale_x_log10()

qplot(data = pf, x = friend_count, binwidth = .1) +
  scale_x_log10()
```

***


### Frequency Polygons

```{r Frequency Polygons}
qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Proportion of users with that friend count',
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(lim = c(300, 1000), breaks = seq(0, 1000, 50))
```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
by(pf$www_likes, pf$gender, sum)
```


***

### Box Plots
Notes:

```{r Box Plots}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot', ylim = c(0, 1000)) 

qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 1000))
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 1000))
```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 250))

by(pf$friend_count, pf$gender, summary)
```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
by(pf$friendships_initiated, pf$gender, summary)
```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}
mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)
sum(pf$mobile_check_in == 1) / length(pf$mobile_check_in)
```

Response:

***

### Analyzing One Variable
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!