Prosper Loan: Exploratory Data Analysis by Kyle Wilbert
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(tidyverse) # all things tidy
library(RColorBrewer) # color palettes
library(lubridate) # help with date manipulation
```

```{r include=FALSE, Load_the_Data}
# Load the Data
loans <- read.csv('prosperLoanData.csv')

# show dimensions of dataset
dim(loans)

# view first few rows to confirm correct load
head(loans)

# view column names
names(loans)

# confirm structure of data
str(loans)

# convert dates using lubridate
loans$LoanOriginationDate <- as_date(loans$LoanOriginationDate)

# create select set of data with fewer variables (21 total variables)
keep_vars <- c("LoanStatus", "LoanOriginationDate", "LoanOriginationQuarter",
               "LoanOriginalAmount", "Term", "ClosedDate",
               "LenderYield", 
               "LoanCurrentDaysDelinquent", 
               "RevolvingCreditBalance", "BorrowerRate", 
               "ProsperRating..Alpha.", 
               "BorrowerState", "EmploymentStatus", 
               "EmploymentStatusDuration", "DebtToIncomeRatio",
               "IncomeRange")
loans2 <-loans[keep_vars]

names(loans2)


```

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
# show dimensions of dataset
dim(loans2)

# reveal structore of loans2
str(loans2)

# summarize loans2
summary(loans2)
```

```{r echo=FALSE}
# view histogram of LoanStatus
summary(loans2$LoanStatus)
qplot(data = loans2, x = LoanStatus)
```
Will need to define Cancelled, Charged-Off, and Defaulted, 
to accurately analyze loss. 
Seems like a fairly large amount of chargeoffs (A charge-off or chargeoff is 
the declaration by a creditor (usually a credit card account) that an amount 
of debt is unlikely to be collected. This occurs when a consumer becomes 
severely delinquent on a debt. Traditionally, creditors will make this 
declaration at the point of six months without payment. 
A charge-off is a form of write-off. 
Source: Wikipedia https://en.wikipedia.org/wiki/Charge-off)
```{r echo=FALSE}
# check order of factors
levels(loans2$LoanOriginationQuarter)
loans2$LoanOriginationQuarter <- factor(loans2$LoanOriginationQuarter, 
                                        levels = 
                                          c("Q4 2005", "Q1 2006", "Q2 2006",
                                            "Q3 2006", "Q4 2006", "Q1 2007",
                                            "Q2 2007", "Q3 2007", "Q4 2007",
                                            "Q1 2008", "Q2 2008", "Q3 2008",
                                            "Q4 2008", "Q2 2009", "Q3 2009", 
                                            "Q4 2009", "Q1 2010", "Q2 2010", 
                                            "Q3 2010", "Q4 2010", "Q1 2011",
                                            "Q2 2011", "Q3 2011", "Q4 2011",
                                            "Q1 2012", "Q2 2012", "Q3 2012",
                                            "Q4 2012", "Q1 2013", "Q2 2013",
                                            "Q3 2013", "Q4 2013", "Q1 2014"))

# view histogram of loans by quarter
qplot(data = loans2, x = LoanOriginationQuarter)

loans2 %>% count(LoanOriginationQuarter)
```
Q1 2009 is missing and Q2 2009 is smaller than Q4 2005 when it appears loans starte to be made originally. 

Source: https://www.lendacademy.com/prosper-review/
```{r echo=FALSE}
# view histogram of LoanOriginalAmount
summary(loans$LoanOriginalAmount)
qplot(data = loans2, x=LoanOriginalAmount, binwidth = 50) +
  scale_x_continuous(limits = c(0,25000), breaks = seq(0, 25000, 2500))
```
Showing the LoanOriginalAmount width a "skinny" binwidth of 50 reveals that
most loans are originated with a round number ending in 5 or 0, as indicated 
by the spikes.

```{r echo=FALSE}
# view histogram of Term
qplot(data = loans2, x = Term, binwidth = 12) +
  scale_x_continuous(limits = c(0,72), breaks = seq(0, 60, 12))
```
The most popular loan term is 36 months (3 years) followed by 
6o months (5 years).
```{r echo=FALSE}
# view histogram of LoanCurrentDaysDelinquent
summary(loans2$LoanCurrentDaysDelinquent)
qplot(data = loans2, x = LoanCurrentDaysDelinquent, binwidth = 15) +
  scale_x_continuous(limits = c(30,3000), breaks = seq(30, 3000, 90))
```

```{r echo=FALSE}
# view histogram of RevolvingCreditBalance
summary(loans2$RevolvingCreditBalance)
qplot(x = RevolvingCreditBalance, data = loans2) +
  scale_x_continuous(limits = c(0,100000), breaks = seq(0,100000,10000))

```

```{r echo=FALSE}
# view histogram of BorrowerRate
summary(loans2$BorrowerRate)
qplot(data = loans2, x = BorrowerRate, binwidth = 0.01)
```
There is somewhat of a binomial distribution - the first between 0.15 
and the other around 0.32. Correlated with DebttoIncome ratio?

```{r echo=FALSE}
# view histogram of BorrowerState
summary(loans2$BorrowerState)

loans2 <- within(loans2, 
                 BorrowerState <- factor(BorrowerState,
                                           levels=names(sort(table(BorrowerState)))))
ggplot(loans2, aes(x = BorrowerState)) +
  geom_bar(stat="count") +
  coord_flip()

# source: https://stackoverflow.com/questions/5208679/order-bars-in-ggplot2-bar-graph?noredirect=1&lq=1
```
The top 5 states originating loans were California, Texas, New York, Florida, 
and Illinois. These states have the largest cities and the largest markets. 
Also, I'd imagine Prosper targeting these markets with marketing campaigns.

```{r echo=FALSE}
# view histogram of EmploymentStatus
qplot(data = loans2, x = EmploymentStatus)
```
The EmploymentStatus variable is problematic. It does not appear that Full-Time, 
Self-Employed, and Part-Time, are a part of the Employed number. Rather than 
being subcategories they were originally coded as separate categories. As-is, 
it's not a very useful variable for measuring risk in this analysis.

```{r echo=FALSE}
# view histogram of EmploymentStatusDuration
summary(loans2$EmploymentStatusDuration)
qplot(data = loans2, x = EmploymentStatusDuration, binwidth = 5)
```
This is an odd variable too, as it could apply to employment or unemployment. 
So we need to look at this variable with EmploymentStatus for this to be useful.

```{r echo=FALSE}
# view histogram of DebtToIncomeRatio
qplot(data = loans2, x = DebtToIncomeRatio, binwidth = 0.1) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  geom_vline(xintercept = 0.43, color = "red")
```
According to the CFBP, 43% is the highest debt-to-income ratio where borrowers 
can still qualify for a mortgage. it appears the bulk of borrowers are below 
that ratio (red line above). 
Source: https://www.consumerfinance.gov/ask-cfpb/what-is-a-debt-to-income-ratio-why-is-the-43-debt-to-income-ratio-important-en-1791/

```{r echo=FALSE}
# reorder factor
loans2$IncomeRange <- factor(loans2$IncomeRange, levels = c("$0", "$1-24,999",
                                                            "$25,000-49,999",
                                                            "$50,000-74,999",
                                                            "$75,000-99,999",
                                                            "$100,000+",
                                                            "Not displayed",
                                                            "Not employed"))

# view histogram of IncomeRange
qplot(data = loans2, x = IncomeRange)

```
The data for income range appears to be a more accurate measure 
of employment and income than the EmploymentStatus variable.

```{r echo=FALSE}
# reorder factors
loans2$ProsperRating..Alpha. <- factor(loans2$ProsperRating..Alpha., levels =
                                          c("AA", "A", "B", "C", "D", "E", "HR"
                                            ))

# view histogram for ProsperRating (Alpha).
qplot(data = loans2, x=ProsperRating..Alpha.)

```
29,084 are missing Alpha ratings.

# Univariate Analysis

### What is the structure of your dataset?
There are 113,937 loans in the modified data set with 16 variables.

Observations
- the majority of loans have a 36 month term
- most loans come from 5 states - CA, TX, NY, FL, IL
- BorrowerRate appears to have a binomial distribution with peaks 
around 15% and 32%
- the majority of borrowers have Debt-to-Income ratios below 43%, 
which is the highest rate allowed to receive a Qualified Mortgage

Anomalies
- ProsperRating..Alpha is missing 29,084 values. 
Need to confirm that they are the same observations
- IncomeRange has both \$0, not employed, and not displayed factors. What's the 
difference between \$0 and not employed? I suppose if you are collecting 
unemployment, your income would not be zero but you would be not employed. 
But what about \$0 income?
- EmploymentStatus is a really messy, unhelpful variable. The counts for 
Full-Time, Part-Time, and Self-Employed are separate from the larger Employed 
category.
- EmploymentStatusDuration is equally as unhelpful because it relies on 
EmploymentStatus. The duration can either be employed/unemployed 
(or any of its subcategories).

### What is/are the main feature(s) of interest in your dataset?
I'm mostly curious about what distinguishes a healthy loan to an unhealthy loan. 
To that end, I will look at LoanStatus, LoanOriginalAmount, 
the Prosper Ratings, BorrowerState, 
DebtToIncomeRatio, and IncomeRange.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
see above

### Did you create any new variables from existing variables in the dataset?
I created Loan OriginationQtr to do some time series analysis

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

- both RevolvingCreditBalance and IncomeToDebtRatio have right-skewed 
distributions. They look exponential.

- otherwise, I tried using some tranforms on certain measures but they didn't
apppear to transform the data much.


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
# view boxplot of BorrowerRate against ProsperRating..Alpha
qplot(x = ProsperRating..Alpha., y= BorrowerRate,
      data = loans2,
      geom = "boxplot")
```
- The borrowers without a ProsperRating have a wide distribution compared to the 
borrowers with a rating - perhaps that's why Prosper developed a rating. 

- There appears to be a positive linear between the best ratings (from best to worst, AA to HR) and lower interest rates.

```{r echo=FALSE}
# BorrowerRate over time
ggplot(loans2) +
  geom_jitter(aes(x = LoanOriginationQuarter, y = BorrowerRate), alpha = 1/100)
```
This shows an interesting picture of the business - pre-2009, there were fewer 
distinct borrower rates. Prior to the gap quarter/reorganization, the 
BorrowerRate widened. After the gap, there remained some high rate loans, but
it appears they stopped doing higher rate loans in 2012. Most recently in 2014,
you can see that they now use finer degrees of interest rates, likely based
on their proprietary credit scoring models.

```{r echo=FALSE}
# calculate upper outliers
summary(loans2$DebtToIncomeRatio) #3Q = 0.32
IQR(loans2$DebtToIncomeRatio, na.rm = TRUE) # IQR = 0.18, 

# view boxplot of DebtToIncomeRatio against PropserRating..Alpha
qplot(data = subset(loans2, DebtToIncomeRatio < 0.59), 
      x = ProsperRating..Alpha., y = DebtToIncomeRatio,
      geom = "boxplot")
```
I removed any outliers in the DebtToIncomeRatio ((0.32) + 1.5(0.18)).
From AA to C we see rising DebtToIncomeRatio as the Prosper Rating falls in
quality. Past the C rating the trend disappears. There does appear to be a
normal distribution of borrowers within each Prosper Rating, by
Debt to Income ratio.

```{r echo=FALSE}
# view boxplot of DebtToIncomeRatio against IncomeRange
qplot(data = subset(loans2, DebtToIncomeRatio < 1),
      x = IncomeRange, y = DebtToIncomeRatio, 
      geom = "boxplot")

```
As you might expect, there is a relationship between DebtToIncomeRatio
and IncomeRange. As income rises, the ratio drops. 

```{r echo=FALSE}
# number of 
by(loans2$LoanCurrentDaysDelinquent,loans2$BorrowerState, length)

# view scatterplot of LoanCurrentDaysDelinquent against BorrowerState
ggplot(aes(x = BorrowerState, y = LoanCurrentDaysDelinquent),
      data = subset(loans2, loans2$LoanCurrentDaysDelinquent > 0)) +
  geom_jitter(alpha = 1/20)
```
There are a surprising number of delinquincies without information about 
origination state (rank:6). It looks like Prosper wasn't collecting enough data or lost
some data at some point.

```{r echo=FALSE}
# summary
by(loans2$EmploymentStatusDuration,loans2$EmploymentStatus, summary) 

# scatterplot of EmploymentStatus vs EmploymentStatusDuration
qplot(x = EmploymentStatus, y = EmploymentStatusDuration,
      data = loans2,
      geom = 'boxplot')
```
Oddly the medians for each category of employment are well less than 1 year 
(96.07 for the category as a whole). Seems odd.
Wonder if this is self-reported or based on credit agency data.

```{r echo=FALSE}
# boxplot of credit balance against rating
qplot(x = ProsperRating..Alpha., y = RevolvingCreditBalance,
      data = loans2,
      geom = 'boxplot') +
  scale_y_continuous(limits = c(0,45000))
  
```
It appears that RevolvingCreditBalance also factors into the Prosper Rating.
As ratings decrease, the median revolving credit also decreases. Also note
that those with an A rating have a slightly higher median than those with a
AA rating.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Covariation
- RevolvingCreditBalance, DebtToIncomeRatio, and BorrowerRate all move parallel
to the Prosper Rating, with BorrowerRate showing the most covariation

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

- shocking number of delinquencies without a BorrowerState recorded
- EmploymentStatus and EmploymentStatusDuration does not appear to show strong
pattern

### What was the strongest relationship you found?
The strongest relationship was between BorrowerRate and ProsperRating

# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}
# view BorrowerRate against LoanOrigination Quarter, color with Rating
ggplot(aes(x = LoanOriginationQuarter, y = BorrowerRate, 
                 color = ProsperRating..Alpha.),
       data = loans2) +
  geom_point() +
  scale_color_brewer(type='qual') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
# Source
# https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2
```
The colors make clear here that pre-2009, there was no Prosper Rating, 
hence no color. After the business rules reset in 2009, we start to see the
implementation of Prosper Ratings. We also see that the better ratings
receive more favorable borrowing rates.

```{r echo=FALSE}
# view BorrowerRate against LoanOriginationQuarter with IncomeRange color
ggplot(aes(x = LoanOriginationQuarter, y = BorrowerRate, 
                 color = IncomeRange),
       data = loans2) +
  geom_point() +
  scale_color_brewer(type='qual') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
Not much of a pattern here - income is not a strong determinant of 
borrower rate. There is some overplotting that I wasn't able to remove without
also removing the color legend.
```{r echo=FALSE}
# x = LoanOriginationQuarter, y = LenderYield, color = ProsperRating..Alpha.

ggplot(aes(x = LoanOriginationQuarter, y = LenderYield, 
                 color = ProsperRating..Alpha.),
       data = loans2) +
  geom_point() +
  scale_color_brewer(type='qual') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
Interesting - this chart looks very similar to the first multivariate chart
we looked at. The difference, however, is that this shows the LenderYield
is higher when they the investor funded riskier loans (based on Prosper Rating).
Basically, there is an inverse relationship between a borrower's Prosper Rating
and the lender's yield. High risk, high reward and vice versa.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

- The ProsperRating helps borrowers secure a lower rate while it is a signal to
lender about which loans will be safe or risky. Safe loans have lower yields 
than more risky loans.

Borrowers benefit from a higher quality rating while lenders see more money
from lower quality ratings.

- Income is not a strong factor in determining borrower rate.
### Were there any interesting or surprising interactions between features?
It's not very surprising when I think about it but seeing the similarity
between charts for borrower rate colored with Rating versus lender yield
colored with Rating was amusing.


### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!